\chapter{Conclusion}
\label{cha:conclusion}
The work done in this thesis centers around Legion, a framework that uses peer-to-peer communication between clients to improve client-to-client latency for collaborative web applications, lowering bandwidth usage on the server and supporting disconnected operations.\par
	Legion supports the integration with legacy storage systems, to offer a safe and stable storage solution and to allow users running new and old applications to interact. This makes possible for client-server applications to run alongside Legion leveraged applications. Starting from this point, our work focused in the integration of Legion with Antidote, one of the goals of the Legion project.\par
	We designed and implemented this integration mechanism, so that an Antidote based application can run alongside with an application that uses Legion. In order to keep the two systems synchronized, we designed and implemented an operation propagation logic, being reactive to events from Legion to Antidote and pro-active pooling from Antidote to Legion.\par
	To validate the implementation, we performed several tests that show us some metrics related to the integration performance and scalability. This metrics are based on message propagation time, stored meta-data size and variables such as number of clients, number of operations and operation message size. In summary, the performed tests showed that propagating data from Legion to Antidote has a higher cost in time per operation than the opposite direction. We also concluded that the message size does not influence operation propagation time. Regarding the metadata stored in Antidote, the size needed scales linearly with the number of operations, being higher with more writes than delete operations.\par
In summary, the main contributions of the work presented in this thesis are as follows:

\begin{itemize}
\item The development of a language independent interface for Antidote.
\item The implementation of a JavaScript client for Antidote.
\item The design and implementation of the integration between Legion and Antidote.
\item The evaluation of the integration mechanism.
\end{itemize}

\section{Technical decisions}
\label{sec:technical_decisions}
In the starting point of this thesis, the main objective was to integrate Legion with two data storage systems that also used CRDTs, Riak and Antidote. Riak is an enterprise product, with several real world use cases, while Antidote is still very fresh and under development. Even thought Antidote uses Riak Core, the two systems diverge a lot based on design options and its objectives. We started by talking to a few developers from each of the systems, being able to conclude that Antidote's development environment would be much more beneficial for us regarding support. This was a critical matter since no one directly involved in this thesis was an expert in either of the systems, neither we had much experience in the programming language used (Erlang). With this in mind, we followed to start the integration with Antidote.\par
	Our first integration design was to use the CRDT's data structures of both systems, translate one to another, and then merge them in order to propagate the state of the system. Once again we talked to the development team to get a guideline of how this could be done. However, it was not the intended design path to expose the internal CRDT's data structure to the outside of the system. The system's CRDTs were meant to only be accessed as an object abstraction.\par
	After acknowledging this, the design of the integration was changed to be based on propagating operations, by executing on one system the operations that were made on the other.

\section{Future Work}
\label{sec:future_work}
After this thesis development, in order to consolidate and complement this integration, a number of future improvements can be mentioned.\par
	The data types available in this implementation are sufficient to demonstrate a working application that uses both these systems, but in a real world scenario, there would be the need to have more data types to correctly support the creation of different application types.\par
	The current Antidote protocol buffer interface allows us send and receive JSON objects serialized as protocol buffers 'bytes'. As this suffices for a working interface, it is not how protocol buffers is intended to be used. Another upgrade, not only to this integration, but also to the Antidote project, would be a refactoring of Antidote's protocol buffer interface.